---
title: "Homework Assignment 2"
subtitle: "Dimensionality Reduction using PCA"
author: "Gunnar Sundberg - gsundberg3513@floridapoly.edu"
output: html_notebook
---


Load the `tidyverse` and `factoextra` packages for this assignment.

```{r, message = FALSE, warning=FALSE}
library(tidyverse)
library(factoextra)
```

# Problem 1

In this problem you will use data from the [EA Sports FIFA 18](https://en.wikipedia.org/wiki/FIFA_18) videogame. 

- Get the dataset `fifa18.csv`: 

```{r, message=FALSE}
fifa18 <- read_csv("https://raw.githubusercontent.com/reisanar/datasets/master/fifa18.csv")
```

- Below is a sample of observations in this dataset:

```{r}
head(fifa18, 8)
```

(a) Use the numerical attributes in this dataset to perform PCA. Get a summary of your PCA results (use the `summary()` function as shown in examples). Comment on your results. 

```{r}
# dataframe with only numerical attributes
myfifa18 <- select(fifa18, -c(1:3))
```

Perform PCA: 

```{r}
fifa18_pca <- prcomp(myfifa18, scale = T)
summary(fifa18_pca)
```

> Comments: In a dataset with 37 usable variables, we are able to capture ~90% of the variation using only the first 9 principal components.



(b) Produce at least one **data visualization** to explain the results on PCA. (You can use the `factoextra` package for this and follow the examples discussed in class)

```{r}
fifa18_pca$rotation[ , 1:9]
```

```{r}
# code here - include as many chunks of code as needed
fviz_screeplot(fifa18_pca)
```
```{r}
fviz_pca(fifa18_pca,
         geom = "point",
         repel = TRUE)
```

> Comments: Using the above visualizations, several things stand out. The biplot shows a clear separation among goalkeeping attributes and all other attributes. This fits with intuition, as goalkeeping relies on a totally separate skillset from those of attackers and defenders. There is also separation between offensive and defensive attributes across the Dim1 axis, with higher Dim2 values being more strongly associated with offensive attributes, and lower values with defensive attributes (not accounting for goalkeaper attributes). Looking at the loadings, we are able to ascertain the most important features in each of the principal components. The first principal component (which explains far and away the most variance, according to our scree plot), positively weighs attributes associated with goalkeeping and negatively weighs those associated with offense. The second weighs positively some attributes associated with attackers and negatively those associated with defenders, just as our biplot analysis showed.

# Problem 2

(a) Read the Toyota Corollas dataset on sales during late summer of 2004 in the Netherlands. It has 1436 records containing details on 38 attributes, including Price, Age, Kilometers, HP, and other specifications. 

```{r, message=FALSE}
toyota <- read_csv("https://raw.githubusercontent.com/reisanar/datasets/master/ToyotaCorolla.csv")
```


- Print the first 6 rows of the dataset.

```{r}
head(toyota)
```


- Identify (to the best of your knowledge) the categorical variables.

> Comments: `model`, `mfg_month`, `mfg_year`, `fuel_type`, `met_color`, `color`, `automatic`, `doors`, `cylinders`, `gears`, `mfr_guarantee`, `bovag_guarantee`, `guarantee_period`, `abs`, `airbag_1`, `airbag_2`, `airco`, `automatic_airco`, `board_computer`, `cd_player`, `central_lock`, `powered_windows`, `power_steering`, `radio`, `mistlamps`, `sport_model`, `backseat_driver`, `metallic_rim`, `radio_cassette`, `parking_assistant`, and `tow_bar` are all categorical variables. This doesn't leave that many numerical variables!


(b) Create a new data frame with some of the numerical attributes in this dataset (ignoring the categorical variables, and think about which variables are probably categorical variables even when represented by a number). 

Perform **Principal Component Analysis (PCA)**, and **comment on your results**.

```{r}
# code here - include as many chunks of code as needed
# you could use the example above to select the collection of variables you want
my_toyota <- toyota %>%
              select(c(price, age_08_04, km, hp, cc, quarterly_tax, weight))
head(my_toyota)
```

```{r}
# code here - include as many chunks of code as needed
my_toyota_pca <- prcomp(my_toyota, scale = T)
summary(my_toyota_pca)
```

> Comments: In a dataset with 7 usable variables, we are able to capture ~90% of the variation using only the first 4 principal components. This is not quite as impressive as the FIFA dataset, but still useful. Let's try to learn more about our data:

```{r}
my_toyota_pca$rotation[ , 1:7]
```

(c) Produce at least one **data visualization** to explain the results on PCA. (You can use the `factoextra` package for this and follow the examples discussed in class)

```{r}
# code here - include as many chunks of code as needed
fviz_screeplot(my_toyota_pca)
```

```{r}
fviz_pca_biplot(my_toyota_pca,
                geom = "point",
                repel = TRUE)
```

> According to the loadings, the first principal component weighs age and mileage positively, with price and weight being negative. The second principal component weighs quarterly tax, mileage, weight, and cc, with horsepower being weighted negatively. Our biplot shows some interesting correlations worth mentioning: age and mileage have a negative correlation with price, weight, cc, and quarterly tax have a positive correlation, and mileage and horsepower are negatively correlated. Do cars with more horsepower tend to not last as long? It could be worth investigating.

## Problem 3

> **Required for graduate students. Extra-credit (+10 points) for undergraduate students)**

You will use data generated by the Spotify API on songs nominated to the _Grammy Award for Song of the Year_.

Find the data in this [GitHub repo](https://github.com/reisanar/datasets/blob/master/grammy_softy_20.csv), and read it into RStudio using:

```{r, message=FALSE}
# read grammy data
soty <- read_csv("https://raw.githubusercontent.com/reisanar/datasets/master/grammy_softy_20.csv")
```

(a) Perform PCA on this dataset, using all the **numerical attributes** available. 


(b) Summarize your findings and comment on your results. 


(c) Apply PCA again to a smaller number of numerical attributes (briefly explain your reasoning for the selection of the numerical attributes you consider for this part of the problem)


(d) Create at least one data visualization of your PCA results.

